---
title: "C°F 5th Weather Chart - Weddings (NL) Analysis"
author: Agathe Lenclen
date: Aug. 2018
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

# C°F 5th Weather Chart - Weddings (NL) Analysis

## About

In this analysis, we compared the number of weddings per week with the temperature and the rain duration.

The weddings data comes from CBS Statline and contains data on population development in the Netherlands, per day. In this table, the data is broken down to the following attributes:

- Births
- Deceased
- Marriage closures (excluding closed partnership registrations)
- Partnership registrations
- (Real) divorces (from 2010 including disbanded partnership registrations)
- Persons moving within the Netherlands
- Foreign migration
- Nationality changes

Data available from: 1995

The steps taken for this analysis are:

- Get weddings data
- Count total number of weddings per day and per year
- Get weather data
- Merge weddings and weather data by day
- Sum the total number of weddings per week for each year
- Average the temperature and rain duration per week over all the years

In the following section, you will find more details about each step.

## Sources 
- Population Development; births, deaths and migration, per day : <http://statline.cbs.nl/Statweb/publication/?DM=SLNL&PA=70703NED&D1=a&D2=8044-8074,l&VW=T>
- KNMI Daily weather : 
    * <http://www.knmi.nl/nederland-nu/klimatologie/daggegevens>
    * <http://www.knmi.nl/kennis-en-datacentrum/achtergrond/data-ophalen-vanuit-een-script>


## Methodology
```{r libs, message=FALSE}

library(plyr)
library(ggplot2)
library(lubridate)
library(dplyr)
source('../utils_NL.R')
```

### Get the marriage data
The CBS Marriage date column is written in plain text, therefore we need to format it to a usable format. 

```{r marriage}
data <- read.csv('../data/NL_Marriages_Divorces_all.csv')
head(data)

data$month <- NA
data$date <- NA

# Format date field
for (i in 1:nrow(data)) {
  splits <- strsplit(as.character(data$Onderwerpen_2[i]), " ")
  length <- length(splits[[1]])

  for (j in 1:length) {
    colname <- paste('split', j)
    value <- splits[[1]][j]
    
    value_date <- as.Date(value, '%d-%m-%Y')

    if (!is.na(value_date)) {
      data[i, 'date'] <- value_date
      break
    } else {
      if (value == 'januari') { data[i, 'month'] <- 1 }
      if (value == 'februari') { data[i, 'month'] <- 2 }
      if (value == 'maart') { data[i, 'month'] <- 3 }
      if (value == 'april') { data[i, 'month'] <- 4 }
      if (value == 'mei') { data[i, 'month'] <- 5 }
      if (value == 'juni') { data[i, 'month'] <- 6 }
      if (value == 'juli') { data[i, 'month'] <- 7 }
      if (value == 'augustus') { data[i, 'month'] <- 8 }
      if (value == 'september') { data[i, 'month'] <- 9 }
      if (value == 'oktober') { data[i, 'month'] <- 10 }
      if (value == 'november') { data[i, 'month'] <- 11 }
      if (value == 'december') { data[i, 'month'] <- 12 }
      
      data[i, colname] <- value
    }
  }
}

for (i in 1:nrow(data)) {
  if (!is.na(data[i, 'month'])) {
    newDate <- paste(data[i, 'split 4'], '-', data[i, 'month'], '-', data[i, 'split 2'], sep='')
    data[i, 'date'] <- as.Date(newDate, '%Y-%m-%d')
  }
}

class(data$date) <- "Date"

data <- data[c('date', 'Totaal.huwelijken', 'X.Echt.scheidingen')]
colnames(data) <- c('date' ,'marriages', 'divorces')
data$year <- year(data$date)
head(data)
```


### Get the weather
For more information on how the weather is treated, see the related reference. 
The marriages data is not geolocated. This data is about the whole Netherlands. Therefore, we have only one option which is to average the weather per day across the whole country.

```{r weather-stations, cache=TRUE}
# KNMI weather stations
nl_stations <- read.csv('../data/all_station_KNMI_cleaned.csv', header=TRUE, sep=",")

# Weather data between 1995 and 2015
weather_data <- getKNMIDailyData('1995', '01', '01', '2015', '12', '31')

# Merge with station data
weather_data_stn <- merge(weather_data, nl_stations, by='STN')

# Format weather data
weather_data_format <- formatNLWeatherDailyData(weather_data_stn)

# Average per day
weather_data_avg <- ddply(weather_data_format, .(date), summarize,  meanFG_kmh=mean(FG, na.rm=TRUE), meanRH_mm=mean(RH, na.rm=TRUE), meanDR_h=mean(DR, na.rm=TRUE), meanTG_C=mean(TG, na.rm=TRUE), meanTX_C=mean(TX, na.rm=TRUE), meanTN_C=mean(TN, na.rm=TRUE), meanSQ_h=mean(SQ, na.rm=TRUE))

head(weather_data_avg)
```


### Merge marriages and weather

```{r merge}
data_weather <- merge(data, weather_data_avg, by='date')
head(data_weather)
```

### Data per week
First let's sum the number of marriages per year and week. Then we'll average the min. and max. temperature per week. 

```{r week1}
sumbyweek <- aggregate(cbind(data_weather$marriages, data_weather$meanRH_mm, data_weather$meanDR_h), by=list(year(data_weather$date), week(data_weather$date)), FUN=sum)
colnames(sumbyweek) <- c('year', 'week', 'sumMarriages', 'sumRH', 'sumDR')
head(sumbyweek)

meanbyweek <- aggregate(cbind(data_weather$meanTX_C, data_weather$meanTN_C), by=list(week(data_weather$date)), FUN=mean)
colnames(meanbyweek) <- c('week', 'meanTX', 'meanTN')
head(meanbyweek)

avgDR <- aggregate(cbind(sumbyweek$sumDR), by=list(sumbyweek$week), FUN=mean)
colnames(avgDR) <- c('week', 'avgDR')
head(avgDR)
```

## Plots

```{r plot}
p1 <- ggplot(data=sumbyweek , aes(x=week, y=sumMarriages)) + geom_line(aes(colour=as.character(year)))+
    xlab("Weeks")+
    ylab("Number of marriages")+
    ggtitle("Number of marriages per week between 1995 and 2015")
p1

p2 <- ggplot(meanbyweek) + geom_line(aes(x=week, y=meanTX, color='tmax')) + geom_line(aes(x=week, y=meanTN, color='tmin'))+
    xlab("Weeks") +
    ylab("Temperature (C)") +
    ggtitle("Average min. and max. temperature per week between 1995 and 2015")
p2

p3 <- ggplot(avgDR, aes(week)) + geom_bar(aes(weight=avgDR)) +
    xlab("Weeks") +
    ylab("Avg. Rain duration per week (hours)") +
    ggtitle("Average rain duration per week between 1995 and 2015")
p3
```
