---
title: "C°F 5th Weather Chart - Mountain Biking (NL, Eindhoven) Analysis"
author: Agathe Lenclen
date: Aug. 2018
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

# C°F 5th Weather Chart - Mountain Biking (NL, Eindhoven) Analysis

## On Strava

Strava is a website and mobile app used to track athletic activity via GPS. One type of data from Strava is called a *Segment* and is defined as "specific sections of road" on which "athletes' times are compared". An athlete performance is called an *Effort*. Back when the analysis was made, Strava API had an available endpoint `/segments/{segment_id}/all_efforts` which allowed to request for all the performance of athletes on a given segment, for a given date range, without the need of athlete's approval. This endpoint is unfortunately not available anymore, thus fetching the data is not reproducible.

## About

In this analysis, we compared the amount of people mountain biking for sport with temperature and rainfall. 

The steps taken for this analysis are:

- Get the number of mountain bike rides per day
- Merge with the weather data
- Aggregate to the total number of rides per week, mean temperature and rainfall per week.

In the following section, you will find more details about each step.

## Sources 
- Strava API : <https://strava.github.io/api/>
- KNMI Daily weather : 
    * <http://www.knmi.nl/nederland-nu/klimatologie/daggegevens>
    * <http://www.knmi.nl/kennis-en-datacentrum/achtergrond/data-ophalen-vanuit-een-script>
- Highlights :
    * http://www.mountainbike.nl/forum/viewtopic.php?f=5&t=118369
    * http://www.wtccw.nl/nieuws/18-aankondiging-activiteit/1229-rondje-geldrop-13-december-2015
    * http://nos.nl/artikel/2014977-knmi-code-oranje-vanwege-ijzel.html


## Methodology

```{r libs, message=FALSE}

library(plyr)
library(ggplot2)
library(lubridate)
library(dplyr)
source('../utils_NL.R')
```

### Find a segment
There are many segments available. After logging in the Strava website, it is possible to [explore segments](https://www.strava.com/segments/explore) for a given location.

The criterias to pick a segment to analyse were:

- The segment should have a large amount of athlete efforts recorded
- The segment should not be a particular segment where the attendance levels could be influenced by a major factor.
For example, a segment located on a university campus could be biased by scheduled sport trainings, and therefore would not reflect a weather influence.

Following this two rules we picked a segment called ["Gulberg Correct"](https://www.strava.com/segments/8596410). It is a mountain biking ride segment in Mierlo, Noord-Brabant, near Eindhoven.


### Get efforts for a segment
By querying the API with the segment ID, a start timestamp and an end timestamp, we get back the list of efforts conducted on the segment. 
For the segment we picked, the ID was `8596410`, and the date range was `start_date_local=2015-01-01T00:00:00Z` to `end_date_local=2015-12-31T23:59:59Z`.
Example of the response head:

```{r query-api,  cache=TRUE}
segmentData <- read.csv('../data/segment_8596410.csv')

# Format date
segmentData$YYYYMMDD <- as.Date(segmentData$start_date_local, "%Y-%m-%d")

head(segmentData)
```

### Get number of mountain bike rides per day
```{r count}
sf_data <- ddply( segmentData , .(YYYYMMDD) , summarise , Count = length(YYYYMMDD), totalDistance = sum(distance), totalTime = sum(elapsed_time), meanTime= mean(elapsed_time), meanDistance=mean(distance) )
names(sf_data)[names(sf_data) == 'YYYYMMDD'] <- 'date'

head(sf_data)
``` 


### Get the weather
For more information on how the weather is treated, see the related reference. 
First we subset the weather stations to those around Eindhoven. For this particular case the subset correspond to one station only.

```{r weather-stations}
# KNMI weather stations
nl_stations <- read.csv('../data/all_station_KNMI_cleaned.csv', header=TRUE, sep=",")
# Get stations of area
area_stations <- nl_stations[ nl_stations$LAT > 51.384754 & nl_stations$LAT < 51.496760 & nl_stations$LON < 5.559295 & nl_stations$LON > 5.2, ]
area_stations
```

Then we get the weather from KNMI and merge it with the station in order to get the weather of that station.
After some formatting the weather data is ready to use.

```{r weather, cache=TRUE}
# Weather data between 2013 and 2015
weather_data <- getKNMIDailyData('2013', '01', '01', '2015', '12', '31')
# Merge with station data
weather_data_stn <- merge(weather_data, area_stations, by='STN')
# Format weather data
weather_data_format <- formatNLWeatherDailyData(weather_data_stn)
head(weather_data_format)
```

### Merge mountain bike rides and weather data
We merge by date.

```{r merge}
sf_data_weather <- merge(sf_data, weather_data_format, by='date', all.y=TRUE)
sf_data_weather$Count[is.na(sf_data_weather$Count)] <- 0

head(sf_data_weather)
```


### Aggregate per week

```{r byweek}
sumbyweek <- aggregate(cbind(Count, RH, SQ)~week(date)+year(date),data=sf_data_weather, FUN=sum, na.rm=TRUE)
colnames(sumbyweek) <- c('week', 'year', 'sumcount', 'sumRH', 'sumSQ')
sumbyweek$year <- as.factor(sumbyweek$year)
head(sumbyweek)

meanbyweek <- aggregate(cbind(Count, TX, TN)~week(date),data=sf_data_weather, FUN=mean, na.rm=TRUE)
colnames(meanbyweek) <- c('week', 'sumcount', 'meanTX', 'meanTN')
head(meanbyweek)

avgRH <- aggregate(cbind(sumbyweek$sumRH), data=sumbyweek, FUN=mean, by=list(sumbyweek$week))
colnames(avgRH) <- c('week', 'avgRH')
head(avgRH)
```

## Plots
```{r plot, fig.width=12, fig.height=8}
p1 <- ggplot(sumbyweek, aes(x=week, y=sumcount)) + geom_line(aes(colour=year)) +
  xlab("Weeks") +
  ylab("Total mountain bike rides") +
  ggtitle("Total mountain bike rides per week and year")

p1

p2 <- ggplot(meanbyweek, aes(x=week)) +
  geom_line(colour='red', aes(y=meanTX)) +
  geom_line(colour='blue', aes(y=meanTN)) +
  xlab("Weeks") +
  ylab("Avg. temperature per week") +
  ggtitle("Avg. min and max temperature per week over three years")

p2

p3 <- ggplot(avgRH, aes(week)) +
  geom_bar(aes(weight=avgRH)) +
  xlab("Weeks") +
  ylab("Total rainfall in a week (mm)") +
  ggtitle("Avg. total rainfall per week")

p3
```

