---
title: "C°F 5th Weather Chart - Weather (NL) Analysis"
author: Agathe Lenclen
date: Aug. 2018
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

# C°F 5th Weather Chart - Weather (NL) Analysis

## About 
The Royal Netherlands Meteorological Institute (KNMI) is the Dutch national weather service. It is possible to retrieve hourly and daily weather data for a given date range and a list of desired weather stations. 

The treatment of the weather data is similar accross the various analysis we made:

- Request hourly or daily data from KNMI endpoints
- Merge with weather station data to have geolocation information, and potentially filter out stations that are not in the area of the analysis
- Format the date field and convert the weather variables fields to the desired units
- Aggregate per time variable (day, week, year) and/or aggregate per weather variable (temperature, rain etc.)

In the following section you will find more details for each step.


## Sources
- KNMI Hourly weather : <http://www.knmi.nl/nederland-nu/klimatologie/uurgegevens>
- KNMI Daily weather :  <http://www.knmi.nl/nederland-nu/klimatologie/daggegevens>
- KNMI Retrieving data by script documentation : <http://www.knmi.nl/kennis-en-datacentrum/achtergrond/data-ophalen-vanuit-een-script>
- KNMI weather code definitions :
    * Daily <http://projects.knmi.nl/klimatologie/daggegevens/selectie.cgi>
    * Hourly <http://projects.knmi.nl/klimatologie/uurgegevens/selectie.cgi>
- KNMI Weather stations data is included in the daily and hourly weather requests.


## Methodology


```{r libs, message=FALSE}

library(plyr)
library(lubridate)
library(sp)
library(rgeos)
library(magrittr)
```

### Get weather data
First let's take a look at a raw file: we will use as an example the dataset from 2015. The URL for the data request can be constructed with a list of weather stations code and a date range. By default we always request all the stations available. The date range is defined as a parameter in a function called `getKNMIDailyData`.

```{r data, cache=TRUE}

# Retrieve daily weather data from KNMI
getKNMIDailyData <- function(startYear, startMonth, startDay, endYear, endMonth, endDay) {
  
  # Construct URL
  url0 <- 'http://projects.knmi.nl/klimatologie/daggegevens/getdata_dag.cgi?stns='
  url1 <- '&vars='
  url2 <- '&byear='
  url3 <- '&bmonth='
  url4 <- '&bday='
  url5 <- '&eyear='
  url6 <- '&emonth='
  url7 <- '&eday='
  
  
  url <- paste(url0, 'ALL', url1, 'all', url2, startYear, url3, startMonth, url4, startDay, url5, endYear, url6, endMonth, url7, endDay, sep="")
  print(url)
  
  # Fetch data
  data <- read.csv(url, header=FALSE, sep=",", comment.char="#", na.strings=" ")
  
  # Format column names
  colnames(data) <- c('STN','YYYYMMDD','DDVEC','FHVEC','FG','FHX','FHXH','FHN','FHNH','FXX','FXXH','TG','TN','TNH','TX','TXH','T10N','T10NH','SQ','SP','Q','DR','RH','RHX','RHXH','EV24','PG','PX','PXH','PN','PNH','VVN','VVNH','VVX','VVXH','NG','UG','UX','UXH','UN','UNH')
  
  return(data)
 
}

# Example
weather_data <- getKNMIDailyData('2015', '01', '01', '2015', '12', '30')
head(weather_data)
nrow(weather_data)
```


The function to request hourly data is similar, except that we need to add the starting and ending hour. Minimum starting hour is 1 and maximum is 24.

```{r hourly, cache=TRUE}

# Retrieve hourly weather data from KNMI
getKNMIHourlyData <- function(startDate, endDate) {
  
  # Construct URL
  url0 <- 'http://projects.knmi.nl/klimatologie/uurgegevens/getdata_uur.cgi?stns='
  url1 <- '&vars='
  url2 <- '&start='
  url3 <- '&end'
  
  url <- paste(url0, 'ALL', url1, 'all', url2, startDate, url3, endDate, sep="")
  print(url)
  
  # Fetch data
  data <- read.csv(url, header=FALSE, sep=",", comment.char="#", na.strings=" ")
  
  # Format column names
  colnames(data) <- c('STN', 'YYYYMMDD',   'HH',   'DD',   'FH',   'FF',   'FX',    'T',  'T10',   'TD',   'SQ',    'Q',   'DR',   'RH',    'P',   'VV',    'N',    'U',   'WW',   'IX',    'M',    'R',    'S',    'O',    'Y')
  
  return(data)
  
}

# Example
weather_hour_data <- getKNMIHourlyData('2015010101', '2015123024')
head(weather_hour_data)
```


### Merge with stations data
Each record correspond to a date and a specific station. 
The list of weather stations with their metadata (latitude, longitude, name, altitude) is available in the header of the file returned by KNMI data service. We manually copied it from an example file and cleaned into a usable CSV. The station code `STN` is the field that will allow us to merge the weather and the station data together.

```{r station, cache=TRUE}
nl_stations <- read.csv('../data/all_station_KNMI_cleaned.csv', header=TRUE, sep=",")
nl_stations

weather_stn <- merge(weather_data, nl_stations, by='STN')
head(weather_stn)
```


### Format weather data
The weather data provided contains all the weather variables available. We can select the type of variables we are interested in, and convert their units: for example the rainfall is given in 0.1mm and we will convert it to 1mm.

```{r format, cache=TRUE}

# Subset weather variables and convert their units
formatNLWeatherDailyData <- function(weather) {
  
  # Date
  weather$date <- as.Date(as.character(weather$YYYYMMDD), "%Y%m%d")
  
  # Select weather variables needed
  weather <- weather[c('STN', 'date', 'FG', 'FHN', 'FHX', 'TG', 'TN', 'TX', 'SQ', 'DR', 'RH', 'LON', 'LAT')]
  
  # Convert units (m/s to km/h, 0.1mm to mm, and 0.1C to C)
  weather$FG <- weather$FG * 0.1 * 3.6
  weather$FHN <- weather$FHN * 0.1 * 3.6
  weather$FHX <- weather$FHX * 0.1 * 3.6
  
  weather$TG <- weather$TG * 0.1
  weather$TN <- weather$TN * 0.1
  weather$TX <- weather$TX * 0.1
  
  weather$SQ <- weather$SQ * 0.1
  
  weather$DR <- weather$DR * 0.1
  weather$RH <- weather$RH * 0.1
  
  weather$RH[weather$RH < 0] <- 0
  
  # Rounded data for potential categorization
  weather$roundedFG <- floor(weather$FG)
  weather$roundedFHN <- floor(weather$FHN)
  weather$roundedFHX <- floor(weather$FHX)
  
  weather$roundedTG <- floor(weather$TG)
  weather$roundedTN <- floor(weather$TN)
  weather$roundedTX <- floor(weather$TX)
  
  weather$roundedSQ <- floor(weather$SQ)
  
  weather$roundedDR <- floor(weather$DR)
  weather$roundedRH <- floor(weather$RH)
  
  return(weather)
  
}

weather_format <- formatNLWeatherDailyData(weather_stn)
head(weather_format)
```

### Get closest weather station to a location

When we needed to find the closest weather station to a given location, we used the following function:

```{r closest}

# Get closest station to location
# Location and StationLocation MUST have their longitude column named `LON` and their latitude column named `LAT`
getClosestWeatherStation <- function(location, stationLocation) {
  
  # Geocode locations
  sp.location <- location
  coordinates(sp.location) <- ~LON+LAT
  
  sp.stations <- stationLocation
  coordinates(sp.stations) <- ~LON+LAT
  
  # Get distance between all points
  d <- gDistance(sp.stations, sp.location, byid=TRUE)
  
  # Get minimum distance to point and merge it with corresponding weather station
  min.d <- apply(d, 1, function(x) order(x, decreasing=F)[1])
  closest_stations <- cbind(location, stationLocation[min.d,], apply(d, 1, function(x) sort(x, decreasing=F)[1]))
  
  return(closest_stations)
  
}

```

This function requires the packages `sp` and `rgeos`.

