---
title: "C°F 5th Weather Chart - Weddings, (US, Travis County, Texas) Analysis"
author: Agathe Lenclen
date: Aug. 2018
output:
  md_document:
    variant: markdown_github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```


# C°F 5th Weather Chart - Weddings, (US, Travis County, Texas) Analysis

## About
In this analysis, we compared the number of people getting married with temperature. 

The Travis County Clerk Web Search presents a list of marriage records with the date of the marriage. 
From that list which can be exported, we can count the number of marriage records per day. 

The steps taken for this analysis are:

- Get weddings data
- Count the number of weddings per day
- Merge with the weather data
- Average the temperature per week over the 5 years
- Count the total number of weddings per week and per year

In the following section you will find more details for each step.

## Sources 
- Travis County Clerk Web Search, Marriage Records : <http://www.tccsearch.org/>
- Highlights :
    * <https://en.wikipedia.org/wiki/Same-sex_marriage_in_Texas>
- NOAA Weather : <ftp://ftp.ncdc.noaa.gov/pub/data/ghcn/daily/by_year/>


## Methodology

``` {r libs, message=FALSE}
source('../utils_US.R')
library(plyr)
library(ggplot2)
library(lubridate)
library(dplyr)
```

### Get the marriages
The marriages analysis uses 5 years of data: the start date is 2011-01-01, the end date is 2015-12-31. 
We extracted the *Date of marriage* column. This is how the list looks like so far:

```{r one-year, echo=FALSE}
marriage_2011 <- read.csv('../data/us_weddings/2011.csv', header = FALSE)
head(marriage_2011)
```

Then we aggregate the dataset by date, by counting the number of times a date is present. Finally we sum the number of marriages per day over a week. For this we used the *week*[^1] function from *lubridate* package.

```{r get-marriages, echo=FALSE}

getMarriagesPerWeek <- function(year){
  
  path <- paste('../data/us_weddings/', year, '.csv', sep='')
  
  marriages <- read.csv(path, header = FALSE)
  colnames(marriages) <- c('date')
  marriages$date <- as.Date(as.character(marriages$date), "%m/%d/%Y")
  
  marriagesPerDay <- aggregate(x=marriages$date, by=list(marriages$date), FUN=length)
  colnames(marriagesPerDay) <- c('date', 'count')
  
  marriagesPerWeek <-  aggregate(cbind(count)~week(date),data=marriagesPerDay,FUN=sum)
  names(marriagesPerWeek)[names(marriagesPerWeek) == 'week(date)'] <- 'WEEK'
  marriagesPerWeek$YEAR <- year
  
  return(marriagesPerWeek)
}

marriages_2015 <- getMarriagesPerWeek('2015')
marriages_2014 <- getMarriagesPerWeek('2014')
marriages_2013 <- getMarriagesPerWeek('2013')
marriages_2012 <- getMarriagesPerWeek('2012')
marriages_2011 <- getMarriagesPerWeek('2011')

marriages <- rbind(marriages_2011, marriages_2012, marriages_2013, marriages_2014, marriages_2015)
head(marriages)
```

### Get the weather data
For more details on how the weather data was treated, see the related reference. 
As we are considering a county, we decided to average the weather data of the weather stations of the county, by date between 2011-01-01 and 2015-12-31. Then we averaged the weather data by week over the 5 years, which gives us a data point per week and year.

``` {r weather, echo=FALSE}
weather_data <- read.csv('../data/travis_weather_data.csv')
head(weather_data)
```

### Marriages and weather
Merging the marriages and weather data by week and year gives us a final dataset :

```{r marriage-weather}
df <- merge(marriages, weather_data, by=c('YEAR', 'WEEK'))
head(df)
```

### Temperature
Let's average the temperature per week over the 5 years:

```{r temperature}
temp <- weather_data %>% group_by(WEEK) %>% summarise(meanTMIN = mean(TMIN), meanTMAX = mean(TMAX))
head(temp)
```

## Plots
```{r plots}
p1 <- ggplot(df, aes(x=WEEK, y=count, color=YEAR)) + geom_line() +
    ggtitle("Total marriages per week") + 
    ylab("Total marriages")
p1

p3 <- ggplot(temp) + geom_line(aes(x=WEEK, y=meanTMAX, color='tmax')) + geom_line(aes(x=WEEK, y=meanTMIN, color='tmin'))+
    ggtitle("Mean min. and max. temperature per week averaged over 5 years") +
    scale_colour_manual(name="Line Color",
    values=c(tmax="red", tmin="steelblue")) +
    ylab("Temperature (C)")
p3

```

## Notes
- On the final graph, we removed the data point for week 53, as it only contains one day. 
- On the final graph, temperature is plotted in Fahrenheit

[^1]: "week returns the number of complete seven day periods that have occured between the date and
January 1st, plus one." <https://cran.r-project.org/web/packages/lubridate/lubridate.pdf>