---
title: "CC°FF 5th Weather Chart - Wildfires (NL, Amsterdam) Analysis"
author: Agathe Lenclen
date: Aug. 2018
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

# C°F 5th Weather Chart - Wildfires (NL, Amsterdam) Analysis

## About

In this analysis, we compared the amount of wildfires with sunshine duration and rainfall. The dataset is an overview of all the fire reports of the Amsterdam-Amstelland Fire Department in the period from 2010 to 2015.

The steps taken for this analysis are:

- Get wildfires data
- Get weather data
- Merge weather data with wildfires data
- Remove the 31st of December and 1st of January data point for each year
- Average the total of wildfires per sunshine duration
- Average the total of rainfall per sunshine duration

In the following section, you will find more details about each step.


## Sources
- Fire Reports 2010-2015 in Amsterdam-Amstelland : <https://data.amsterdam.nl/index.html#?dte=dcatd%2Fdatasets%2Fbrandmeldingen-2010-2015&dtfs=T&mpb=topografie&mpz=9&mpv=52.3719:4.9012>
- KNMI Daily weather :
    * <http://www.knmi.nl/nederland-nu/klimatologie/daggegevens>
    * <http://www.knmi.nl/kennis-en-datacentrum/achtergrond/data-ophalen-vanuit-een-script>
- Information on fires : <https://nl.wikipedia.org/wiki/Brand_(vuur)>

## Methodology

```{r libs, message=FALSE}
# Load libraries
library(plyr)
library(ggplot2)
library(lubridate)
library(dplyr)

# Load util functions
source('../utils_NL.R')
```

### Get fires data
The fires dataset contains the date of the fires reports, the type of incident and the location. There are more fields available but they are not needed for the analysis.

```{r sr, cache=TRUE}
data <- read.csv('../data/brwaa_2010-2015.csv', header=TRUE, sep=";")

# Format date column
data$date <- as.Date(data$start_tijd, "%Y/%m/%d")

# Remove trailing whitespace
data$incident_type <- trimws(data$incident_type)
data$gemeente <- trimws(data$gemeente)

head(data)
```

### Get wildfires data in Amsterdam area

```{r agg}
# Get data for amsterdam only
data_ams <- data[ data$gemeente == "Amsterdam", ]

# Get the list of incident types
incident_types <- unique(data_ams[c("incident_type")])
incident_types

# Filter to get only wildfires (Buitenbrand) incidents
data_ams_out <- data_ams[data_ams$incident_type == as.character(incident_types[2, 'incident_type']), ]

# Get total of wildfires per day
sum_data <- aggregate(x=data_ams_out$date, by=list(data_ams_out$date), FUN=length)
colnames(sum_data) <- c('date', 'Count')
head(sum_data)
```

### Get the weather
For more information on how the weather is treated, see the related reference.
For this analysis we will use the weather data of Schipol weather station (STN=240). We get the weather data from KNMI and subset it to the desired station. After some formatting the weather data is ready to use.

```{r weather, cache=TRUE}
# KNMI weather stations
nl_stations <- read.csv('../data/all_station_KNMI_cleaned.csv', header=TRUE, sep=",")
# Weather data between 2005 and 2015
weather_data <- getKNMIDailyData('2005', '03', '31', '2015', '05', '20')
# Merge with station data
weather_data_stn <- merge(weather_data, nl_stations, by='STN')
# Subset for station 240
weather_data_240 <- weather_data_stn[ weather_data_stn$STN == 240, ]
# Format weather data
weather_data_format <- formatNLWeatherDailyData(weather_data_240)
head(weather_data_format)
```


### Merge weather and wildfires data

```{r merge}
sum_data_weather <- merge(sum_data, weather_data_format, by='date', all.y=TRUE)

# For days without wildfires, set count to zero
sum_data_weather$Count[is.na(sum_data_weather$Count)] <- 0
```

### Remove New Year's Eve outliers
New Year's Eve comes with a huge peak of wildfires each year, which skew all the rest of the data. We will isolate the 31st of December and the 1st of January of each year.

```{r nye-plot}
ggplot(sum_data_weather, aes(x=date, y=Count)) + geom_line()
```

```{r nye}
df_outnye <- sum_data_weather[ !(month(sum_data_weather$date) == 1 & day(sum_data_weather$date) == 1) & !(month(sum_data_weather$date) == 12 & day(sum_data_weather$date) == 31), ]
```

### Aggregate per sunshine duration
Sunshine duration is given in hours by the variable SQ. We will use the column roundedSQ which is the rounded value of the original SQ value, to an integer.

```{r sq}
df_sq <- df_outnye %>% group_by(roundedSQ) %>% summarise(meanCount=mean(Count, na.rm=TRUE), meanRH = mean(RH, na.rm=TRUE))
df_sq
```

## Plots

```{r plot}
p1 <- ggplot(df_sq, aes(roundedSQ)) + geom_bar(aes(weight=meanCount)) +
    ylab("Average number of wildfires per day") +
    xlab("Sunshine duration per day (hours)") +
    ggtitle("Avg. number of wildfires per day depending on the sunshine duration")
p1

p2 <- ggplot(df_sq, aes(roundedSQ)) + geom_bar(aes(weight=meanRH)) +
    ylab("Average rainfall per day (mm)") +
    xlab("Sunshine duration per day (hours)") +
    ggtitle("Avg. rainfall per day depending on the sunshine duration")
p2
```
