---
title: "C°F 5th Weather Chart - Cycling, (US, Palo Alto) Analysis"
author: Agathe Lenclen
date: Aug. 2018
output:
  md_document:
    variant: markdown_github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```


# C°F 5th Weather Chart - Cycling, (US, Palo Alto) Analysis

## On Strava

Strava is a website and mobile app used to track athletic activity via GPS. One type of data from Strava is called a *Segment* and is defined as "specific sections of road" on which "athletes' times are compared". An athlete performance is called an *Effort*. Back when the analysis was made, Strava API had an available endpoint `/segments/{segment_id}/all_efforts` which allowed to request for all the performance of athletes on a given segment, for a given date range, without the need of athlete's approval. This endpoint is unfortunately not available anymore, thus fetching the data is not reproducible.


## About

In this analysis, we compared the number of people cycling with temperature and rainfall. 

The steps taken for this analysis are:

- Get the number of cycling rides per day
- Merge with the weather data
- Aggregate to the total number of rides per week, mean temperature and cumulative rainfall per week.

In the following section, you will find more details about each step.

## Sources 
- Strava API : <https://strava.github.io/api/>
- Highlights :
    * <http://sanfrancisco.cbslocal.com/2015/05/14/south-bay-hail-lightning-and-thunder-storm-rain-drought/>
    * <https://www.ncdc.noaa.gov/stormevents/eventdetails.jsp?id=554921>
- NOAA Weather : <ftp://ftp.ncdc.noaa.gov/pub/data/ghcn/daily/by_year/>


## Methodology

``` {r libs, message=FALSE}
source('../utils_US.R')
library(plyr)
library(ggplot2)
library(lubridate)
```


### Find a segment
There are many segments available. After logging in the Strava website, it is possible to [explore segments](https://www.strava.com/segments/explore) for a given location.

The criterias to pick a segment to analyse were:

- The segment should have a large amount of athlete efforts recorded
- The segment should not be a particular segment where the attendance levels could be influenced by a major factor.
For example, a segment located on a university campus could be biased by scheduled sport trainings, and therefore would not reflect a weather influence.

Following this two rules we picked a segment called ["Old La Honda (Bridge to Mailboxes)"](https://www.strava.com/segments/8109834). It is a ride segment in Woodside, California.


### Get efforts for a segment
By querying the API with the segment ID, a start timestamp and an end timestamp, we get back the list of efforts conducted on the segment. 
For the segment we picked, the ID was `8109834`, and the date range was `start_date_local=2015-01-01T00:00:00Z` to `end_date_local=2015-12-31T23:59:59Z`.
Example of the response head:
```{r query-api,  echo=FALSE}
segmentData <- read.csv('../data/segmentData_8109834.csv')
head(segmentData)
```


### Get the number of rides per day
The goal is to obtain the amount of rides per day. The number of records for a given date is what gives us this amount. 
```{r aggregate}
# Format date
segmentData$YYYYMMDD <- as.Date(segmentData$start_date_local, "%Y-%m-%d")

# Aggregate by date
sf_data <- ddply(segmentData , .(YYYYMMDD) , summarise , Count = length(YYYYMMDD))
head(sf_data)
```

### Get the weather data
For more details on how the weather data was treated, see the related reference. 
As we are considering a 4km cycling path, we decided to average the weather data of the closest weather stations around the cycling path, which accounts to 22 weather stations.
The weather data and the segment data are then merged by date.

```{r get-weather, echo=FALSE}
weather_sf_average <- read.csv('../data/weather_sf_average_2015.csv')
# Format date
weather_sf_average$date <- as.Date(weather_sf_average$date , "%Y-%m-%d")
```

```{r merge-weather}
# Change aggregated segment data  column name to date
names(sf_data)[names(sf_data) == 'YYYYMMDD'] <- 'date'

# Merge aggregated segment data with the weather data
sf_data_weather <- merge(sf_data, weather_sf_average, by='date')
head(sf_data_weather)
```

### Aggregate data by week
Then we aggregate the data per week, as it is easier to read the chart this way. For this we used the *week*[^3] function from *lubridate* package.


```{r byweek}
meanbyweek <- aggregate(cbind(Count, PRCP, TMAX, TMIN)~week(date),data=sf_data_weather,FUN=mean)
colnames(meanbyweek) <- c("week", "meanCount", "meanPRCP", "meanTMAX", "meanTMIN")
sumbyweek <- aggregate(cbind(Count, PRCP, TMAX, TMIN)~week(date),data=sf_data_weather,FUN=sum)
sumbyweek <- sumbyweek[c("week(date)", "Count", "PRCP")]
colnames(sumbyweek) <- c("week", "totalCount", "totalPRCP")
```

## Plots

```{r plot}
l <- rep(1, nrow(sumbyweek))
p1 <- ggplot(sumbyweek, aes(x=week, y=totalCount)) + geom_line() + ggtitle("Total amount of rides per week")
p1
p2 <- ggplot(sumbyweek, aes(x=week, y=l, size=totalPRCP)) + geom_point() + ggtitle("Cumulative rainfall per week (mm)") + theme(axis.line=element_blank(),
        axis.text.y=element_blank(),
        axis.title.y=element_blank()) +
        scale_size_area()
p2
p3 <- ggplot(meanbyweek) + geom_line(aes(x=week, y=meanTMAX, color='tmax')) + geom_line(aes(x=week, y=meanTMIN, color='tmin'))+
    ggtitle("Mean min. and max. temperature per week") +
    scale_colour_manual(name="Line Color",
    values=c(tmax="red", tmin="steelblue")) +
    ylab("Temperature (C)")
p3
```


## Notes
- On the final graph, we show only rainfall values that are superior to 0.1 inches per week
- On the final graph, we removed the data point for week 53, as it only contains one day. 
- On the final graph, temperature is plotted in Fahrenheit and rainfall is plotted in inches. 


[^1]: https://en.wikipedia.org/wiki/Strava.
[^2]: https://strava.github.io/api/v3/segments/
[^3]: "week returns the number of complete seven day periods that have occured between the date and
January 1st, plus one." https://cran.r-project.org/web/packages/lubridate/lubridate.pdf